### 2018-10-11 write-up/dissertation/evaluation.tex
Organise write-ups

@@

\begin{landscape}
\begin{figure}[p]
    \centering
    \begin{minipage}{1.45\textheight}
    \begin{minted}[linenos, fontsize=\small]{c}
static void kalman( const int n,               const int k,                const double *sigma, /* n,n */
                    const double *h, /* k,n */ const double *mu, /* n,1 */ double *r,           /* k,k */
                    double *data,    /* k,1 */ double **ret_mu,  /* k,1 */ double **ret_sigma   /* n,n */ ) {
        double* k_by_n = (double *) malloc(k * n * sizeof(double));
/*16*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
/*17*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, k, k, n, 1., k_by_n, n, h, n, 1., r, k);
/*18*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
/*19*/  cblas_dcopy(k * n, h, 1, k_by_n, 1);
        double* k_by_k = (double *) malloc(k * k * sizeof(double));
/*20*/  cblas_dcopy(k * k, r, 1, k_by_k, 1);
/*21*/  LAPACKE_dposv(LAPACK_ROW_MAJOR, 'U', k, n, k_by_k, k, k_by_n, n);
/*23*/  LAPACKE_dpotrs(LAPACK_ROW_MAJOR, 'U', k, 1, k_by_k, k, data, 1);
        free(k_by_k);
        double* n_by_n = (double *) malloc(n * n * sizeof(double));
/*24*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, n, k, 1., h, n, k_by_n, n, 0., n_by_n, n);
        free(k_by_n);
        double* n_by_1 = (double *) malloc(n * sizeof(double));
/*25*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, 1, k, 1., h, n, data, 1, 0., n_by_1, 1);
        double* new_mu = (double *) malloc(n * sizeof(double));
/*26*/  cblas_dcopy(n, mu, 1, new_mu, 1);
/*27*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, 1, 1., sigma, n, n_by_1, 1, 1., new_mu, 1);
        free(n_by_1);
        double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
/*28*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, 1., sigma, n, n_by_n, n, 0., n_by_n2, n);
/*29*/  cblas_dcopy(n*n, sigma, 1, n_by_n, 1);
/*30*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, n, -1., sigma, n, n_by_n2, n, 1., n_by_n, n);
        free(n_by_n2);
        *ret_sigma = n_by_n;
        *ret_mu = new_mu; }
    \end{minted}
    \caption{CBLAS/LAPACKE implementation of a Kalman filter. I used C instead
        of Fortran because it is what Owl uses under the hood and OCaml FFI
        support for C is better and easier to use than that for Fortran. A distinct
        `measure\_kalman' function that sandwiches a call to this function with
        \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}

    \end{minipage}
\end{figure}
\end{landscape}

@@

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let potrs ~uplo a b =
  let b = Owl.Mat.copy b in
  Owl.Lapacke.potrs ~uplo ~a ~b
;;

>>>>

\begin{landscape}
\begin{figure}[p]
    \centering
    \begin{minted}[fontsize=\footnotesize]{c}
static void kalman( const int n,               const int k,                const double *sigma, /* n,n */
                    const double *h, /* k,n */ const double *mu, /* n,1 */ double *r,           /* k,k */
                    double *data,    /* k,1 */ double **ret_mu,  /* k,1 */ double **ret_sigma   /* n,n */ ) {
        double* k_by_n = (double *) malloc(k * n * sizeof(double));
/*16*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, k, n, 1., sigma, n, h, n, 0., k_by_n, n);
/*17*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, k, k, n, 1., k_by_n, n, h, n, 1., r, k);
/*18*/  cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
/*19*/  cblas_dcopy(k * n, h, 1, k_by_n, 1);
        double* k_by_k = (double *) malloc(k * k * sizeof(double));
/*20*/  cblas_dcopy(k * k, r, 1, k_by_k, 1);
/*21*/  LAPACKE_dposv(LAPACK_ROW_MAJOR, 'U', k, n, k_by_k, k, k_by_n, n);
/*23*/  LAPACKE_dpotrs(LAPACK_ROW_MAJOR, 'U', k, 1, k_by_k, k, data, 1);
        free(k_by_k);
        double* n_by_n = (double *) malloc(n * n * sizeof(double));
/*24*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, n, k, 1., h, n, k_by_n, n, 0., n_by_n, n);
        free(k_by_n);
        double* n_by_1 = (double *) malloc(n * sizeof(double));
/*25*/  cblas_dgemm(CblasRowMajor, CblasTrans, CblasNoTrans, n, 1, k, 1., h, n, data, 1, 0., n_by_1, 1);
        double* new_mu = (double *) malloc(n * sizeof(double));
/*26*/  cblas_dcopy(n, mu, 1, new_mu, 1);
/*27*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, 1, 1., sigma, n, n_by_1, 1, 1., new_mu, 1);
        free(n_by_1);
        double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
/*28*/  cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, 1., sigma, n, n_by_n, n, 0., n_by_n2, n);
/*29*/  cblas_dcopy(n*n, sigma, 1, n_by_n, 1);
/*30*/  cblas_dsymm(CblasRowMajor, CblasLeft, CblasUpper, n, n, -1., sigma, n, n_by_n2, n, 1., n_by_n, n);
        free(n_by_n2);
        *ret_sigma = n_by_n;
        *ret_mu = new_mu; }
    \end{minted}
    \caption{CBLAS/LAPACKE implementation of a Kalman filter. I used C instead
        of Fortran because it is what Owl uses under the hood and OCaml FFI
        support for C is better and easier to use than that for Fortran. A distinct
        `measure\_kalman' function that sandwiches a call to this function with
        \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}

\end{figure}
\end{landscape}

@@

@@

\begin{figure}[tp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let potrs ~uplo a b =
  let b = Owl.Mat.copy b in
  Owl.Lapacke.potrs ~uplo ~a ~b
;;

### 2018-10-11 write-up/dissertation/introduction.tex
Organise write-ups

Linear types allow the compiler and programmer to statically keep track of the
resources that an implementation uses, thus offering a promising solution to
resource management. However, they have not made their way into mainstream
programming languages. To demonstrate their usefulness, I implemented an OCaml
library that allows users to learn about, use and become familiar with linear
types, specifically in the context of linear-algebra programs.

@@

    \item Many \textbf{new} examples of how the type system can \textbf{automatically
        check} for common aliasing, read/write permissions, memory allocation,
        re-use and deallocation \textbf{errors} in the context of
        linear-algebra programs.

    \item An \textbf{original} demonstration of how that type system can be
        \textbf{useful} for \textbf{matrix expression compilation}.

    \item \textbf{New and readable implementations} of \textbf{non-trivial}
        linear-algebra programs that \textbf{take advantage} of said type
        system.

    \item A \textbf{new solution} to the \textbf{dichotomy} of
        \textbf{readability}, \textbf{ease of reasoning and safety} of
        high-level linear-algebra libraries versus the
        \textbf{memory-efficiency} of low-level linear-algebra libraries.

    \item A \textbf{new library design} to provide said solution in a way that
        \textbf{integrates well} with existing OCaml code and linear-algebra
        libraries.

>>>>

Linear types allow the compiler and programmer to statically keep track of the
resources that a program uses, thus offering a promising solution to the
problems associated with complex resource management. However, they have not
made their way into many mainstream programming languages, in the same way
parametrically polymorphic types have. To illustrate their simplicity and
power, I implemented an OCaml library that allows users to learn about and
become familiar with linear types, specifically in the context of linear
algebra programs.

@@

    \item Many \textbf{new} examples of how that type system can \textbf{automatically
        check} for common aliasing, read/write permission, memory allocation,
        re-use and deallocation \textbf{errors} in the context of
        linear algebra programs.

    \item An \textbf{original} demonstration of how that type system can be
        \textbf{used} for \textbf{matrix expression compilation}.

    \item \textbf{New and readable implementations} of \textbf{non-trivial}
        linear algebra programs that \textbf{take advantage} of said type
        system.

    \item A \textbf{new solution} to the \textbf{dichotomy} of
        \textbf{readability}, \textbf{ease of reasoning and safety} of
        high-level linear algebra libraries versus the
        \textbf{memory-efficiency} of low-level linear algebra libraries.

    \item A \textbf{new library design} to provide said solution in a way that
        \textbf{integrates well} with existing OCaml code and linear algebra
        libraries.

### 2018-10-11 write-up/dissertation/related_work.tex
Organise write-ups

Clak and Cl1ck~\cite{fabregat_thesis} were developed independently around the
same time as SymPy's matrix expression compilation. Clak attempts to produce
\emph{multiple} algorithms for a single matrix expression, by considering a
wider matrix expression grammar and more matrix properties and inference rules.
These algorithms assume basic building blocks such as products and
factorisations. Cl1ck attempts to take on the challenge of writing BLAS/LAPACK
like libraries too, by generating lower-level loop-based blocked routines for
the aforementioned basic building blocks, in the spirit of the
FLAME~\cite{gunnels_flame} project.

@@

In LT4LA, I took the approach of having a concrete syntax and expression
language which was then translated and made available as a \emph{typed
expression} to other modules \emph{at compile time} via the build system. Apart
from convenience in programming and testing, there was nothing inherent in the
approach that prevented me from using OCaml's PPX syntax-extensions so that I
could write normal OCaml expressions from within OCaml and have them checked
for linearity before compilation.

@@

They perform rudimentary pattern-matching and in some cases, loop-fusion, to
avoid evaluating expressions in a purely binary manner (invoking the bane of a
C++ programmer: temporaries and unnecessary copies) when possible (either by
translating to a library kernel call or, as an example, inlining a $v := a + b
+ c$ vector expression into one loop).

As is the case with LT4LA, this approach shares \emph{some} elimination of
runtime overheads, but not all, thanks to the heuristics surrounding lazy
evaluation and evaluating sub-expressions into temporaries. This comes at the
cost of a user being able to easily inspect the generated C++ code, losing
explicitness.

@@

Apart from lazy evaluation, the following projects show how instead of a (E)DSL
library approach, we could have type-level resource management provided far
more conveniently and naturally at the \emph{language} level. The difference is
that a library can be designed, shipped and used now whereas language features
take time and can have unintended interactions with other language features. My
hope is that once people are convinced of the utility type-level resource
management by using a library, the impetus for integrating such features into
the language follows.

@@

A particularly strong advantage LT4LA has over other libraries that use lazy
evaluation is, funnily enough, linear types; more precisely, it is the static
and perfect information they guarantee about aliasing: with Owl, every
graph-node has only one node in or out; with Eigen, every ``assignment'' has a
\texttt{noalias} annotation, without the danger of getting it wrong and
invoking undefined behaviour.

This simplifies the rules and exceptions a programmer reasoning about memory
usage needs to remember. Of course, now the programmer has to figure out how
they are using their temporaries, but because matrices are linearly-typed,
redundant copies/missed frees can be pointed out by the compiler, guiding them
towards a satisfying solution.

@@

Futhark~\cite{futhark} is a second-order (meaning it supports functions such as
map and fold/reduce) array combinator (meaning array operations can be fused
into streams to reduce temporaries) language designed for efficient parallel
compilation. It supports ML-style modules, loops, limited parametric
polymorphism, size types and uniqueness types (an idea closely related to
linear types). Its combinators are more expressive than typical linear-algebra
library kernels so encourage shorter, more declarative linear-algebra code.

@@

Rust~\cite{rust} is a (relatively) new systems programming language aiming to
bring the last two decades of programming language research to the masses in a
usable and friendly manner. Its \emph{borrow-checker} is the feature most
relevant to this thesis because it statically attempts to prevent many
resource-related bugs. Although there are a few linear-algebra libraries for
Rust under development, careful use of its macro system and borrow-checker
could make it the safest and easiest to use language for linear-algebra
projects to come. Its struggle is more likely to be against the inertia of the
large amounts of C++/Fortran code already out there rather than its usability
or benefits. Given this inertia, I believe there is value in taking the DSL
library approach of LT4LA which will work with existing systems.

@@

Dimension mismatches are seen as an irritating but small inconvenience when
writing linear-algebra code. However, another error I found in the Fortran code
output by SymPy to compute a Kalman filter was a dimension/transposition error.
Although we would not need full dependent types to solve dimension mismatches
(symbolic size types would be sufficient), managing properties about matrices
could be done at the type-level in a dependently typed setting.

>>>>

Clak and Cl1ck~\cite{fabregat_thesis} were developed independently around the
same time as SymPy's matrix expression compilation. Clak attempts to produce
\emph{multiple} algorithms for a single matrix expression, by considering a
wider matrix expression grammar and more matrix properties and inference rules.
These algorithms assume basic building blocks such as products and
factorisations. Cl1ck attempts to take on the challenge of writing
BLAS/LAPACK-like libraries too, by generating low-level loop-based blocked
routines for the aforementioned basic building blocks, in the spirit of the
FLAME~\cite{gunnels_flame} project.

@@

In LT4LA, I took the approach of having a concrete syntax and expression
language which was then translated to OCaml and made available to other modules
\emph{at compile time} via the build system. There was nothing inherent in the
approach that prevented me from using OCaml's PPX syntax-extensions so that I
could write normal OCaml expressions from within OCaml and have them checked
for linearity before compilation.

@@

Libraries that use expression templates usually perform rudimentary pattern-matching and in some cases, loop-fusion, to
avoid evaluating expressions in a purely binary manner (preventing the bane of a
C++ programmer: temporaries and unnecessary copies) when possible (either by
translating to a library kernel call or, as an example, inlining a $v := a + b
+ c$ vector expression into one loop).

Therefore, expression templates eliminate some runtime overheads, but not all
(unlike LT4LA), because the heuristics they use rely on statically imperfect
information. They also do not allow a user to easily inspect the generated code
(also unlike LT4LA), thus losing explicitness.

@@

Apart from lazy evaluation, the following projects show how instead of a (E)DSL
library approach, we could have type-level resource management provided far
more conveniently and naturally at the \emph{language} level. The difference is
that a library can be designed, shipped and used now whereas language features
take time and can have unintended interactions with other language features. My
hope is that once people are convinced of the utility type-level resource
management by using a library, the impetus for integrating such features into
a language follows.

@@

A particularly strong advantage LT4LA has over other libraries that use lazy
evaluation is, funnily enough, linear types; more precisely, it is the static
and perfect information they guarantee about aliasing: with Owl, this would be
comparable to assuming every vertex of its computation graph has only one edge
in and out; with Eigen, this would be comparable to having a \texttt{noalias}
annotation on every ``assignment'', \textbf{without the danger of getting it
wrong} and invoking undefined behaviour.

This simplifies the rules and exceptions a programmer reasoning about memory
usage needs to remember. Of course, now the programmer has to figure out how
they are using their temporaries, but because matrices are linearly typed,
redundant copies and missed frees can be pointed out by the compiler, thus
guiding the programmer towards a satisfying solution.

@@

Futhark~\cite{futhark} is a second-order (meaning it supports functions such as
map and fold/reduce) array combinator (meaning array operations can be fused
into streams to reduce temporaries) language designed for efficient parallel
compilation. It supports ML-style modules, loops, limited parametric
polymorphism, size types and uniqueness types (an idea closely related to
linear types). Its combinators are more expressive than typical linear algebra
library kernels; hence it encourages shorter, more declarative code.

@@

Rust~\cite{rust} is a (relatively) new systems programming language aiming to
bring the last two decades of programming language research to the masses in a
usable and friendly manner. Its \emph{borrow checker} is the feature most
relevant to this thesis because it also statically attempts to prevent many
resource-related bugs. There are a few linear algebra libraries for
Rust under development; careful use of its macro system and borrow checker
could make it the safest and easiest to use language for linear algebra
projects to come. Its struggle is more likely to be against the inertia of the
large amounts of C, C++ and Fortran code already out there rather than its
usability or benefits. Given this inertia, I believe there is value in taking
the DSL library approach of LT4LA, which will work with existing systems.

@@

Dimension mismatches are seen as an irritating but small inconvenience when
writing linear algebra code. However, another error I found in the Fortran code
output by SymPy to compute a Kalman filter was a dimension/transposition error.
Although we would not need full dependent types to solve dimension mismatches
(symbolic size types would be sufficient), managing properties about matrices
could be done at the type-level in a dependently typed setting.
